---
title: "Practical 3: Decomposition of AirPassengers Time Series Data"
author: "Praveer Raj"
date: "2026-02-07"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(stats)
```

# Practical 4: Decomposition of AirPassengers Time Series Data

**Praveer Raj**  
**Roll:** 1  
**Reg. No.:** 230957002

---

## 1. Title

Consider the following datasets from the R library and write R code to decompose the 
time-series data into its trend, seasonal, and residual components. Furthermore, identify 
the dominating component(s) in the dataset:  

(b) AirPassengers data
---

## 2. Objective

To load the AirPassengers dataset, convert it into a time series object, visualize passenger trends over time, and decompose the data into trend, seasonal, and irregular components.

---

## 3. R Code

```{r code, results='hide', fig.show='hide'}
# Clear workspace
rm(list = ls())

# Load stats package
library(stats)

# Load AirPassengers dataset
data(AirPassengers)

# Convert to time series object
air_ts <- ts(AirPassengers, start = c(1949, 1), frequency = 12)

# Plot original data
plot(air_ts,
     main = "AirPassengers Time Series",
     xlab = "Year",
     ylab = "Number of Passengers")

# Decompose the time series
air_decomp <- decompose(air_ts)

# Plot decomposed components
plot(air_decomp)

# Display decomposition values
air_decomp
```

---

## 4. Output

### Original Time Series Plot

```{r output-plot1, echo=FALSE}
plot(air_ts,
     main = "AirPassengers Time Series",
     xlab = "Year",
     ylab = "Number of Passengers")
```

### Decomposed Time Series Components

```{r output-plot2, echo=FALSE}
plot(air_decomp)
```

### Decomposition Values

```{r output-values, echo=FALSE}
air_decomp
```

---

## 5. Conclusion

The AirPassengers dataset shows a strong upward trend in the number of airline passengers over time. The decomposition reveals a pronounced seasonal pattern, with recurring peaks each year. The trend component indicates long-term growth in air travel, while the irregular component represents random fluctuations. This analysis demonstrates that the dataset contains both trend and seasonal characteristics.

