---
title: "Practical 3: Decomposition of Nottem Time Series Data"
author: "Praveer Raj"
date: "2026-02-07"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(stats)
```

# Practical 3: Decomposition of Nottem Time Series Data

**Praveer Raj**  
**Roll:** 1  
**Reg. No.:** 230957002

---

## 1. Title

Consider the following datasets from the R library and write R code to decompose the 
time-series data into its trend, seasonal, and residual components. Furthermore, identify 
the dominating component(s) in the dataset:  

(a) Nottem data 


---

## 2. Objective

To load the *nottem* dataset, convert it into a time series object, visualize the original data, and decompose it into its trend, seasonal, and irregular components.

---

## 3. R Code

```{r code, results='hide', fig.show='hide'}
# Clear workspace
rm(list = ls())

# Load stats package
library(stats)

# Load nottem dataset
data(nottem)

# Convert to time series object
nottem_ts <- ts(nottem, start = c(1920, 1), frequency = 12)

# Plot original data
plot(nottem_ts,
     main = "Nottem Monthly Temperature Data",
     xlab = "Year",
     ylab = "Temperature")

# Decompose the time series
nottem_decomp <- decompose(nottem_ts)

# Plot decomposed components
plot(nottem_decomp)

# Display decomposition values
nottem_decomp
```

---

## 4. Output

### Original Time Series Plot

```{r output-plot1, echo=FALSE}
plot(nottem_ts,
     main = "Nottem Monthly Temperature Data",
     xlab = "Year",
     ylab = "Temperature")
```

### Decomposed Time Series Components

```{r output-plot2, echo=FALSE}
plot(nottem_decomp)
```

### Decomposition Values

```{r output-values, echo=FALSE}
nottem_decomp
```

---

## 5. Conclusion

The decomposition of the *nottem* dataset separates the time series into trend, seasonal, and irregular components. A strong seasonal pattern is observed, reflecting recurring monthly temperature variations. The trend component shows gradual long-term changes, while the irregular component captures random fluctuations. This decomposition helps in understanding the structure of the time series data.

