---
title: "Practical 4: Trend Analysis of US Population Time Series"
author: "Praveer Raj"
date: "2026-02-07"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(forecast)
```

# Practical 4: Trend Analysis of US Population Time Series

**Praveer Raj**  
**Roll:** 1  
**Reg. No.:** 230957002

---

## 1. Title

For a given dataset of the US Population (in millions), write the R code to: 
(a) Import the data as a time series object. 
(b) Identify the dominating component(s) in the data set. 
(c) Apply a square root transformation to the original data set. 
(d) Estimate the linear trend present in the data set. 
(e) Remove the estimated linear trend from the data set. 
 
 
 
 
 
Year Population 
1970 3929214 
1971 5308483 
1972 7239881 
1973 9638453 
1974 12860702 
1975 17063353 
1976 23191876 
1977 31443321 
1978 38558371 
1979 50189209 
1980 62979666 
1981 76212168 
1982 92228496 
1983 106021537 
1984 123202624 
1985 132164569 
1986 151325798 
1987 179323175 
1988 203302031 
1989 226542203 
1990 248709873
---

## 2. Objective

To analyze the US population time series by applying transformation, estimating the trend using linear regression, and examining the detrended series.

---

## 3. R Code

```{r code, results='hide', fig.show='hide'}
# Clear workspace
rm(list = ls())

library(forecast)

year <- 1970:1990
population <- c(
  3929214, 5308483, 7239881, 9638453, 12860702,
  17063353, 23191876, 31443321, 38558371, 50189209,
  62979666, 76212168, 92228496, 106021537, 123202624,
  132164569, 151325798, 179323175, 203302031,
  226542203, 248709873
)

population_ts <- ts(population, start = 1970, frequency = 1)

plot(population_ts,
     main = "US Population Time Series",
     xlab = "Year",
     ylab = "Population")

population_sqrt <- sqrt(population_ts)

plot(population_sqrt,
     main = "Square Root Transformed Population",
     xlab = "Year",
     ylab = "Sqrt(Population)")

time_index <- time(population_ts)
trend_model <- lm(population_ts ~ time_index)

summary(trend_model)

estimated_trend <- trend_model$fitted.values

plot(population_ts,
     main = "US Population with Estimated Trend",
     xlab = "Year",
     ylab = "Population")
lines(estimated_trend, col = "red", lwd = 2)

detrended_population <- population_ts - estimated_trend

plot(detrended_population,
     type = "o",
     main = "Detrended US Population",
     xlab = "Year",
     ylab = "Population (Detrended)")
```

---

## 4. Output

### Original Time Series

```{r output1, echo=FALSE}
plot(population_ts,
     main = "US Population Time Series",
     xlab = "Year",
     ylab = "Population")
```

### Square Root Transformation

```{r output2, echo=FALSE}
plot(population_sqrt,
     main = "Square Root Transformed Population",
     xlab = "Year",
     ylab = "Sqrt(Population)")
```

### Estimated Trend with Regression Line

```{r output3, echo=FALSE}
plot(population_ts,
     main = "US Population with Estimated Trend",
     xlab = "Year",
     ylab = "Population")
lines(estimated_trend, col = "red", lwd = 2)
```

### Detrended Time Series

```{r output4, echo=FALSE}
plot(detrended_population,
     type = "o",
     main = "Detrended US Population",
     xlab = "Year",
     ylab = "Population (Detrended)")
```

### Regression Model Summary

```{r output5, echo=FALSE}
summary(trend_model)
```

---

## 5. Conclusion

The US population time series shows a strong upward linear trend. The square root transformation stabilizes variation in the data. Linear regression effectively estimates the trend component, and detrending removes long-term growth, allowing clearer observation of fluctuations. This analysis demonstrates how transformation and trend estimation help in understanding time series structure.

